---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lacunarity

<!-- badges: start -->
<!-- badges: end -->

The goal of lacunarity is to estimate lacunarity and generalized lacunarity for unidimensional time series.

## Methodology



## Installation

You can install the released version of lacunarity from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("lacunarity")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Ikarobarreto/lacunarity")
```
## Example

We are going to simulate a two sets of 1200 bernoulli random variables of 0's an 1's with probabilities of success of 80% and 50% calculating lacunairty of them using lac function:

```{r example 1}
library(lacunarity)
x<-rbinom(n=1200,size=1,prob=.8)
y<-rbinom(n=1200,size=1,prob=.5)

lac(x)
lac(y)
## basic example code
```

The results showed $$\gamma$$, $$\Delta(s)$$, and possible $$s$$. We can observe that $$s$$ parameter has less possible values for y than x and we could infer is because of lower probability of success. However, this is not true, because some coded time series could have 50% of 0's and has different lacunatiries. For example, lets simulate a 25 sequences of 0's and 1's sub-sequences of size 8 and 16 each: 

```{r example 2}
library(lacunarity)
z<-c(rep(c(rep(1,8),rep(0,8)),25),rep(c(rep(1,16),rep(0,16)),25))
table(z)
lac(x)
lac(z)
## basic example code
```

As we can see, both series x and z has around 50% probability of 0's and 1's but different lacunarities.

Now, we are going to simulate a two sets of 10000 bernoulli random variables of 0's an 1's with probabilities of success of 80% and 50% calculating generalized lacunairty of them using genlac function:

```{r example 3}
library(lacunarity)
x<-rbinom(n=10000,size=1,prob=.8)
y<-rbinom(n=10000,size=1,prob=.5)

genlac(x)
genlac(y)
## basic example code
```

The results showed $$s$$,$$q$$, $$\gamma(s)$$, and $$\Delta_q(s)$$.
