---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(2021)
```

# lacunarity

<!-- badges: start -->
<!-- badges: end -->

**lacunarity** estimates the lacunarity and the generalized lacunarity of
unidimensional binary time series. Lacunarity is a scale-dependent measure of
*translational heterogeneity*: it describes how the gaps of a pattern are
distributed, and — unlike density alone — it tells apart series that are
clustered from series that are evenly spread.

## Methodology

For a sliding box of size $s$, the box mass $m$ is the number of ones it covers.
From the distribution of box masses, the lacunarity index is

$$\Lambda(s) = \frac{\langle m^2 \rangle}{\langle m \rangle^2} = 1 + \frac{\mathrm{Var}(m)}{\langle m \rangle^2} \ge 1,$$

so $\Lambda(s) = 1$ means a perfectly homogeneous pattern and larger values mean
gappier, more heterogeneous textures. The *generalized* lacunarity
$\Lambda_q(s) = \left(Z_{2q}(s)/Z_q(s)^2\right)^{1/q}$ extends this to an
arbitrary moment order $q$, yielding a spectrum of scaling exponents
$\gamma(q)$. See `vignette("lacunarity")` for the full theory and references
(Allain & Cloitre, 1991; Vernon-Carter et al., 2009).

## Installation

You can install the released version from
[CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("lacunarity")
```

and the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Ikarobarreto/lacunarity")
```

## Example

`lac()` returns the lacunarity scaling exponent `y`, the lacunarities `Ds` at
each scale and the box scales `s`:

```{r lac}
library(lacunarity)

x <- rbinom(1200, size = 1, prob = 0.8)
lac(x)
```

Lacunarity captures texture, not just density. The series `z` (blocks of ones
and zeros) and `w` (a random shuffle of `z`) have the **same** number of ones,
yet very different lacunarities:

```{r texture}
z <- c(rep(c(rep(1, 8), rep(0, 8)), 25), rep(c(rep(1, 16), rep(0, 16)), 25))
w <- sample(z)

c(density_z = mean(z), density_w = mean(w))      # identical density
c(lac_z = lac(z)$y, lac_w = lac(w)$y)            # different lacunarity
```

`genlac()` returns the generalized lacunarity; its scaling spectrum `yq` over the
moment orders `q` summarises how small and large gaps scale:

```{r genlac}
g <- genlac(x)
round(g$yq, 3)
```

See `vignette("lacunarity")` for plots and a full walkthrough.
